Measurement Bias from Address Aliasing

### 2016-01-09 paper.tex
Place blas graph on correct page
Reintroduce pmu comment and reference

\section{Experimental Methodology}
\label{sec:methodology}
In order to understand biased behavior it is important to be able to do precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
This can be achieved by using performance counters; instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
Performance counters are supported in the Linux kernel, and can be accessed via a tool called \emph{perf}.
This utility instructs the kernel to enable the processor's Performance Monitoring Unit (PMU), before executing a specified program.
Using this technique avoids modifications to the executed program, unlike tools like PAPI~\cite{PAPI:PortableInterface}.
We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.

%Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
%The issues they find are either handled explicitly or are irrelevant to our setup.
%They find intra-machine variations of up to 1.07\%,

A small Python script is used to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
Only a small set of events are collected at a time, to ensure events are actually counted continuously.
We do not use multiplexing between a limited set of counter registers, as it will introduce significant variations~\cite{Wiplove:ImproveEstimation}. % Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
% Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
Results are also averaged over multiple runs to reduce potential random error.%, using built-in perf rerun functionality.

@@

%\newpage
\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
\iffalse
The functionality is divided into three categories:
\begin{description}
  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
\end{description}
Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
\fi
%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
%In code, the function is invoked with the following parameters:
%\begin{lstlisting}[breaklines=true]
%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
%\end{lstlisting}

@@

\pgfplotstableread{bin/libblas.dat}{\libblastable}
\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
\begin{figure*}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8, % Make it fit to text width side by side
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libatlas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libopenblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:blas}Performance of \texttt{cblas\_dgemv} for different BLAS packages, measuring cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$. Each sample point represents an increment of 16 bytes, or 0x10.}
\end{figure*}

>>>>

\section{Experimental Methodology}
\label{sec:methodology}
In order to understand biased behavior it is important to be able to do precise and detailed measurements, without introducing \emph{observer effects}~\cite{Mytkowicz:2008:OE&MB}.
This can be achieved by using performance counters; instrumentation support in hardware that can be used to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
Recent Intel architectures have several hundred available events, providing a detailed view of what happens inside the CPU.
Performance counters are supported in the Linux kernel, and can be accessed via a tool called \emph{perf}.
This utility instructs the kernel to enable the processor's Performance Monitoring Unit (PMU), before executing a specified program.
Using this technique avoids modifications to the executed program, unlike tools like PAPI~\cite{PAPI:PortableInterface}.
We use the perf-stat command in all experiments, which accepts raw PMU event codes listed in the reference manual~\cite{Volume3B}.
Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
The issues they find are either handled explicitly or are irrelevant to our setup.
%They find intra-machine variations of up to 1.07\%,

A small Python script is used to iterate over an exhaustive set of all known counters, which amounts to about 200 on our architecture.
Only a small set of events are collected at a time, to ensure events are actually counted continuously.
We do not use multiplexing between a limited set of counter registers, as it will introduce significant variations~\cite{Wiplove:ImproveEstimation}. % Paper is rename of "Toward Accurate Performance Evaluation using Hardware Counters".
Controlled variations in environment size is performed by setting a dummy environment variable to $n$ number of zero characters.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
% Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
Results are also averaged over multiple runs to reduce potential random error.%, using built-in perf rerun functionality.

@@

%\newpage
\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
\pgfplotstableread{bin/libblas.dat}{\libblastable}
\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
\begin{figure*}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8, % Make it fit to text width side by side
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libatlas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libopenblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:blas}Performance of \texttt{cblas\_dgemv} for different BLAS packages, measuring cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$. Each sample point represents an increment of 16 bytes, or 0x10.}
\end{figure*}

\iffalse
The functionality is divided into three categories:
\begin{description}
  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
\end{description}
Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
\fi
%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
%In code, the function is invoked with the following parameters:
%\begin{lstlisting}[breaklines=true]
%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
%\end{lstlisting}

@@
